\textit{MSF: Modulated Sub-graph Finder}

### 2018-03-20 MSF_Paper_FA.tex
firest few paragraphs

\documentclass[twocolumn]{article}
\usepackage{authblk}
\usepackage{blindtext}
\usepackage{multicol}% http://ctan.org/pkg/multicols
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{cite}
\usepackage[table,xcdraw]{xcolor}
\newcommand{\FA}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\TODO}[1]{\begingroup\color{red}#1\endgroup}
%opening
\begin{document}

@@

\author[1]{Mariam R Farman}
\author[1,2]{Fabian Amman}
\author[1]{Ivo L Hofacker}
\affil[1]{Institute for Theoretical Chemistry,Theoretical Biochemistry Group,University of Vienna,Austria}
\affil[2]{Department of Chromosome Biology, Max F. Perutz Laboratories,University of Vienna,Austria}

\maketitle

\begin{abstract}
Differentially expressed genes play an important role in giving
insights to a phenotype. These DEGs are then used to identify the
pathways altered during changing biological conditions. We developed a
novel approach to find the maximal significantly dis-regulated
sub-graphs or cluster of genes from the cell signaling network, giving
these sub-graphs an overall significance of modulation by combining the
individual \textit{p}-values of the genes.
\end{abstract}

@@

\section*{Background}

High throughput sequencing techniques have been widely used to yield
the differentially expressed genes (DEG)~\cite{DEG}. To this end,
changes in transcript abundance are measured, e.g.~by next generation
sequencing techniques, and interpreted as an indicator of differential
expression of genes. DEGs can be used to get
insights into the mechanism underlying differences between conditions of samples, such as
healthy versus diseased. Differential gene
expression analysis (DGEA) informs about the magnitude of expression changes
between the conditions which are often expressed as fold change, sign of fold change and the confidence level
of observing a authentic change, often expressed as
\textit{p}-value. These DEGs are
then used as information converted from data, this information is further interpreted to knowledge to extract meaningful insights, for example the genes that
could be important for a particular condition or maybe the target gene
of an infection or disease. To this end, pathway-based analysis has become an
important tool to further interpret the results of a DGEA and to
acquire understandings of the perturbations in a biological
system. Biological pathways are sets of genes and their interactions forming a functional
unit. DEGs help to identify pathways or networks that may be altered
during a change of condition providing important information about
diseases and its treatment process~\cite{Khatri2012}. Pathway-based
methods use predefined pathways or networks such as
KEGG~\cite{Kegg} and Reactome~\cite{Reactome}, the expression
measurements of the genes obtained from DGEA in combination with statistical methods and algorithms to
identify specifically modulated pathways and processes~\cite{Campos}.

Predefined pathways are sets of genes and their interactions categorized in a functional unit differently by different databases known as pathways. For example, KEGG (Kyoto Encyclopedia of Genes and Genomes) is a database resource for understanding the functionality and utilities of the biological system. KEGG pathways is a branch of KEGG database that has a collection of manually drawn pathway maps representing the molecular interaction, reaction and relation networks of cellular functions~\cite{Kegg}. Similarly, Reactome is an open-source, manually curated, peer-reviewed and well known database for signaling and metabolic molecules with their interactions formed into different  biological pathways for nineteen species~\cite{Reactome}.

Existing pathway-based analysis approaches use different research
designs, which can be categorized into ORA (Over-representation
analysis), FCS (Functional class scoring) and pathway topology based
methods. ORA is the first and the most basic method of pathway
analysis. It uses a DEG list with user defined cut-off for the log-fold change and
\textit{p}-value (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05). Subsequently, sets of genes associated
with annotated pathways are tested for being over-represented in the
set of DEGs. To this end, hyper-geometric
distribution, chi-square tests, binomial probability or the Fisher’s
exact test are used. Thereby the information of the topology of genes in the
pathways are ignored~\cite{Bayer}. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

The supplementary file contains tables of comparison between \textbf{MSF} identified sub-graph gene enrichment with SPIA and Reactome pathway analysis. 
\clearpage

\begin{thebibliography}{9}

	\bibitem{DEG}
	Malone, J. H., \& Oliver, B. (2011). Microarrays, deep sequencing and the true measure of the transcriptome. BMC Biology, 9(1), 34. doi:10.1186/1741-7007-9-34
	\bibitem{Khatri2012}
	Khatri, P., Sirota, M., \& Butte, A. J. (2012). Ten Years of Pathway Analysis: Current Approaches and Outstanding Challenges. PLoS Computational Biology, 8(2). doi:10.1371/journal.pcbi.1002375
	\bibitem{Kegg}
	Kanehisa, M. (2000). KEGG: Kyoto Encyclopedia of Genes and Genomes. Nucleic Acids Research, 28(1), 27-30. doi:10.1093/nar/28.1.27
	\bibitem{Reactome}
	Croft, D., \& Croft, D. (2010). Reactome: A database of biological pathways. Nature Precedings. doi:10.1038/npre.2010.5025
	\bibitem{Campos}
	García-Campos, M. A., Espinal-Enríquez, J., \& Hernández-Lemus, E. (2015). Pathway Analysis: State of the Art. Frontiers in Physiology, 6. doi:10.3389/fphys.2015.00383 
	\bibitem{Bayer}
	Bayerlová, M., Jung, K., Kramer, F., Klemm, F., Bleckmann, A., \& Beißbarth, T. (2015). Comparative study on gene set and pathway topology-based enrichment methods. BMC Bioinformatics, 16(1). doi:10.1186/s12859-015-0751-5   
	\bibitem{Hartung}
	Hartung, J. (1999). A Note on Combining Dependent Tests of Significance. Biometrical Journal, 41(7), 849-855. doi:10.1002/(sici)1521-4036(199911)41:7<849::aid-bimj849>3.0.co;2-t 5

	\bibitem {Prins}
	Prins, K., Cardenas, W. and Basler, C. (2009). Ebola Virus Protein VP35 Impairs the Function of Interferon Regulatory Factor-Activating Kinases IKK and TBK-1. Journal of Virology, 83(7), pp.3069-3077.

	\bibitem {Konde}
	Konde, M., Baker, D., Traore, F., Sow, M., Camara, A., Barry, A., Mara, D., Barry, A., Cone, M., Kaba, I., Richard, A., Beavogui, A., Günther, S., Pintilie, M. and Fish, E. (2017). Interferon β-1a for the treatment of Ebola virus disease: A historically controlled, single-arm proof-of-concept trial. PLOS ONE, 12(2).

	\bibitem {Rhein}
	Rhein, B., Powers, L., Rogers, K., Anantpadma, M., Singh, B., Sakurai, Y., Bair, T., Miller-Hunt, C., Sinn, P., Davey, R., Monick, M. and Maury, W. (2015). Interferon-γ Inhibits Ebola Virus Infection. PLOS Pathogens, 11(11), p.e1005263.

	\bibitem {Falasca}
	Falasca, L., Agrati, C., Petrosillo, N., Di Caro, A., Capobianchi, M., Ippolito, G. and Piacentini, M. (2015).
	 Molecular mechanisms of Ebola virus pathogenesis: focus on cell death. Cell Death and Differentiation, 22(8), pp.1250-1259.

	\bibitem {Cardenas}
	Cardenas, W., Loo, Y., Gale, M., Hartman, A., Kimberlin, C., Martinez-Sobrido, L., Saphire, E. and Basler, C. (2006). Ebola Virus VP35 Protein Binds Double-Stranded RNA and Inhibits Alpha/Beta Interferon Production Induced by RIG-I Signaling. Journal of Virology, 80(11), pp.5168-5178.

	\bibitem {Olejnik}
	Olejnik, J., Forero, A., Deflubé, L., Hume, A., Manhart, W., Nishida, A., Marzi, A., Katze, M., Ebihara, H., Rasmussen, A. and Mühlberger, E. (2017). Ebolaviruses Associated with Differential Pathogenicity Induce Distinct Host Responses in Human Macrophages. Journal of Virology, 91(11), pp.e00179-17.

	\bibitem {Dilley}
	Dilley, K., Voorhies, A., Luthra, P., Puri, V., Stockwell, T., Lorenzi, H., Basler, C. and Shabman, R. (2017). The Ebola virus VP35 protein binds viral immunostimulatory and host RNAs identified through deep sequencing. PLOS ONE, 12(6), p.e0178717.

	\bibitem{Michalak}
	Michalak, P. (2008). Coexpression, coregulation, and cofunctionality of neighboring genes in eukaryotic genomes. Genomics, 91(3), 243-248. doi:10.1016/j.ygeno.2007.11.002

	\bibitem{Cyto}
	http://www.cytoscape.org/

	\bibitem{Cytokegg}
	http://apps.cytoscape.org/apps/cytokegg

	\bibitem{Reactome}
	https://wiki.reactome.org/index.php?title=ReactomeFIViz

	\bibitem{Tarca}
	Tarca, A. L., Draghici, S., Khatri, P., Hassan, S. S., Mittal, P., Kim, J., … Romero, R. (2009). A novel signaling pathway impact analysis. Bioinformatics, 25(1), 75–82. http://doi.org/10.1093/bioinformatics/btn577

	\bibitem{StringApp}
	http://apps.cytoscape.org/apps/stringapp

	\bibitem{Veljkovic}
	Veljkovic, V., Glisic, S., Muller, C. P., Scotch, M., Branch, D. R., Perovic, V. R., … Colombatti, A. (2015). In silico analysis suggests interaction between Ebola virus and the extracellular matrix. Frontiers in Microbiology, 6, 135. http://doi.org/10.3389/fmicb.2015.00135

	\bibitem{edgeR}
	Robinson, M. D., Mccarthy, D. J., \& Smyth, G. K. (2009). EdgeR: a Bioconductor package for differential expression analysis of digital gene expression data. Bioinformatics, 26(1), 139-140. doi:10.1093/bioinformatics/btp616

\end{thebibliography}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Welcome to Overleaf --- just edit your article on the left,
% and we'll compile it for you on the right. If you give
% someone the link to this page, they can edit at the same
% time. See the help menu above for more info. Enjoy!
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% For more detailed article preparation guidelines, please see:
% http://f1000research.com/author-guidelines

\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage{f1000_styles}

%% Default: numerical citations
\usepackage[numbers]{natbib}

%% Uncomment this lines for superscript citations instead
% \usepackage[super]{natbib}

%% Uncomment these lines for author-year citations instead
% \usepackage[round]{natbib}
% \let\cite\citep

\usepackage{subcaption}
\usepackage{hyperref}
\newcommand{\FA}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\TODO}[1]{\begingroup\color{red}#1\endgroup}
%opening
\begin{document}

@@

\author[1]{Mariam R. Farman}
\author[1,2]{Fabian Amman}
\author[1]{Ivo L. Hofacker}
\affil[1]{Institute for Theoretical Chemistry,Theoretical Biochemistry Group,University of Vienna, Austria}
\affil[2]{Department of Chromosome Biology, Max F. Perutz Laboratories,University of Vienna, Austria}

\maketitle
\thispagestyle{fancy}

\begin{abstract}

High throughput techniques such as RNA-seq or microarray analysis have
proven to be invaluable for the characterization of global
transcriptional gene activity changes due to external stimuli or
diseases. Differential gene expression analysis (DGEA) is the first
step in the course of data interpretation, typically producing lists
of of dozens to thousands of differentially expressed genes. To
further guide the interpretation of these lists, different pathway
analysis approaches have been developed. Most (\TODO{all?}) of which
classify genes into set of genes, i.e. a pathway, which do interact
with each other function in a common biological process. Regardless of
technical differences, all these methods do not properly account for
cross talk between different pathways and rely on binary
classification into differentially expressed gene and unaffected genes
based on an arbitrarily set p-value cut-off.

To overcome this limitation, we developed a novel approach to identify
concertedly modulated sub-graphs in the global cell signaling network,
based on the DGEA results of all genes tested. Thereby, expression
patterns of genes are integrated according to the topology of their
interactions and allow potentially to read the flow of information
from the perturbation source to the effectors. The described software,
names \texttt{Modulated Sub-graph Finder} is freely available at
\url{http//:path/to/git}.

\end{abstract}

@@

\clearpage

\section*{Introduction}

High throughput sequencing techniques have been widely used to yield
differentially expressed genes (DEG)~\cite{DEG}. To this end, changes
in transcript abundance are measured, e.g.~by next generation
sequencing techniques, and interpreted as an indicator of differential
expression of genes. DEGs can be used to get insights into the
mechanism underlying differences between conditions of samples, such
as healthy versus diseased. Differential gene expression analysis
(DGEA) informs about the magnitude of expression changes between the
conditions which are often expressed as fold change, sign of fold
change and the confidence level of observing an authentic change,
often expressed as \textit{p}-value. These DEGs information is further
interpreted to extract meaningful biological insights. For example,
genes that could be involved in the response to a particular stimuli
or maybe the cause of a disease. To this end, pathway-based analysis
has become an important tool to further interpret the results of a
DGEA and to acquire understandings of the perturbations in a
biological system. Biological pathways are sets of genes and their
interactions forming a functional unit. DEGs help to identify pathways
or networks that may be altered during a change of condition providing
important information about diseases and its treatment
process~\cite{Khatri2012}. Pathway-based methods use predefined
pathways or networks such as KEGG~\cite{Kegg} and
Reactome~\cite{Reactome}, the expression measurements of the genes
obtained from DGEA in combination with statistical methods and
algorithms to identify specifically modulated pathways and
processes~\cite{Campos}.

Well established resources for pathway annotation are KEGG (Kyoto
Encyclopedia of Genes and Genomes)\cite{Kegg} and
Reactome~\cite{Reactome}. KEGG pathways is a branch of KEGG database
that has a collection of manually drawn pathway maps representing the
molecular interaction, reaction and relation networks of cellular
functions. Similarly, Reactome is an open-source, manually curated,
peer-reviewed database for signaling and metabolic molecules with
their interactions formed into different biological pathways for
nineteen species~\cite{Reactome}. Both provide predefined pathways
which are sets of genes and their interactions categorized into
functional units. Starting from a gene interaction network, genes are
labeled according their role in a specific biological process. In this
sense a particular gene can be assigned to different pathways. E.g.,
the human gene STAT1 is associated with 24 different pathways in the
pathway annotation curated by
KEGG~\footnote{\url{http://www.genome.jp/dbget-bin/www_bget?hsa:6772}}
and in 12 different pathways in the Reactome data
base~\footnote{\url{http://www.reactome.org/content/detail/R-HSA-629622}}.
Although carefully produced, the assignment of genes to those
predefined pathway unites can be considered to be subjective to some
degree and suffers from observatinal bias (\TODO{citation needed}).

\TODO{FA until here}.

Existing pathway-based analysis approaches use different research
designs, which can be categorized into ORA (Over-representation
analysis), FCS (Functional class scoring) and pathway topology based
methods. All of which aim in finding in a subset of genes, e.g.,
significantly differentially expressed genes, genes associated with a
certain pathway more often than expected given the total set of
examined genes, e.g. the genome background.  ORA is the first and the
most basic method of pathway analysis. It uses a DEG list with user
defined cut-off for the log-fold change and \textit{p}-value (most
commonly using absolute log-fold change $\geq$ 2 and \textit{p}-value
$\leq$ 0.05). Subsequently, sets of genes associated with annotated
pathways are tested for being over-represented in the set of DEGs. To
this end, hyper-geometric distribution, chi-square tests, binomial
probability or the Fisher’s exact test are used. Thereby the
information of the topology of genes in the pathways are
ignored~\cite{Bayer}. Furthermore, ORA assumes that the biological
pathways are independent of each other and ignores the fact that they
cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

The supplementary file contains tables of comparison between
\textbf{MSF} identified sub-graph gene enrichment with SPIA and
Reactome pathway analysis.

{\small\bibliographystyle{unsrtnat}
\bibliography{MSF_Paper_FA}}

\bigskip
